Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-24 blinded.Rmd
ready version

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
bibliography: references.bib 
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
library(doParallel)
library(doRNG)
library(foreach)
library(doFuture)
```

@@

We choose the ARMA(3,2), as out simple benchmark model. The corresponding log-likelihood value comes out to be **-105.93**.

@@

bird_aic_table <- aic_table(data$logCPUE,4,5)
require(knitr)
kable(bird_aic_table,digits=2,caption = "AIC Table.")
```

@@

bird_ts <- ts(bird$logCPUE)
arma_model <- Arima(bird_ts, order = c(3, 0, 2))

@@

@@

```{r, echo=FALSE, eval=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out
```

A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions. Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model. This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as to control the robustness. The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.

```{r, echo=FALSE, eval=FALSE}
bake(file="local_search.rds", {
  foreach(i=1,.combine=c,
    .options.future=list(seed=482947940)
  ) %dofuture% {
    mod |>
      mif2(
        Np=1000, Nmif=50,
        cooling.fraction.50=0.5,
        rw.sd = rw_sd(
           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
           a = 0.02, b = 0.02, c = 0.02, 
          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
        ),
        partrans = parameter_trans(
             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')),
        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs"),
        params = param_guess
      )
  }}) -> mifs_local

@@

```{r, message=FALSE}
# bake(file="local_search.rds", {
#   foreach(i=1:20,.combine=c,
#     .options.future=list(seed=482947940)
#   ) %dofuture% {
#     mod |>
#       mif2(
#         Np=1000, Nmif=50,
#         cooling.fraction.50=0.5,
#         rw.sd = rw_sd(
#            alpha = 0.02, Beta = 0.02, gamma = 0.02,
#            a = 0.02, b = 0.02, c = 0.02,
#           sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
#          sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
#         ),
#         partrans = parameter_trans(
#              log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
#               'c', 'logF_0', 'logB_0', 'logRho')),
#         paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB",
#                 "logF_0", "logB_0", "logRho", "sigma_obs"),
#         params = param_guess
#       )
#   }}) -> mifs_local

mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y")
```

The likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

```{r}
# bake(file="lik_local.rds",{
#   foreach(mf=mifs_local,.combine=rbind,
#     .options.future=list(seed=900242057)
#   ) %dofuture% {
#     evals <- replicate(10, logLik(pfilter(mf,Np=100)))
#     ll <- logmeanexp(evals,se=TRUE)
#     mf |> coef() |> bind_rows() |>
#       bind_cols(loglik=ll[1],loglik.se=ll[2])
#   } -> results
#   attr(results,"ncpu") <- nbrOfWorkers()
#   results
# }) -> results_local
# bind_rows(results_local) |> arrange(-loglik) |> write_csv("bird_params.csv")
# if (file.exists("CLUSTER.R")) {
#   source("CLUSTER.R")
# }
pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```

@@

@@

# Results

Our report probes the potential of applying POMP to the field of ecological population dynamics. We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models. Drawing inspiration from the Lotka-Voltera equations, we introduce our take on the concept of an alternative prey hypothesis. Although our analysis is on the side of preliminary, it shows this approach can shed more light on the dynamic responses of ecosystems to changes in prey abundance.

# Contribution

- Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
- Heyuan: 
- Sizhuang:

>>>>

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Blinded"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
bibliography: references.bib 
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
plan(multisession)
library(doParallel)
library(doRNG)
library(foreach)
library(doFuture)
```

@@

ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for a stationarity check.
```{r stationary, echo=FALSE}
# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss_result <- tseries::kpss.test(ts_cpue, null = "Level")
print(kpss_result)

# Differencing the time series once
start_year <- min(data$year)
end_year <- max(data$year)
frequency <- 1  
ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
ts_diff <- diff(ts_cpue, differences = 1)
print(tseries::kpss.test(ts_diff,null = "Level"))
```

KPSS with p-value = 0.01 suggests that the average daily hunting of birds(CPUE) in log scale is non-stationary. However, first-order difference on this time series data is stationary, proved by the p-value larger than $\alpha=0.05$ from KPSS test. Then the following ARMA(p,q) model will be run on such differenced data.

@@

bird_aic_table <- aic_table(ts_diff,4,5)
require(knitr)
kable(bird_aic_table,digits=2,caption = "AIC Table.")
```

The best ARMA(0,5) is selected given the lowest AIC with 204.48 to fit first-order differenced data. Thus, ARIMA(0,1,5) will be run below to check for its log-likelihood.

@@

bird_ts <- ts(bird$logCPUE)
arma_model <- Arima(bird_ts, order = c(0, 1, 5))

@@

We choose the ARIMA(0,1,5), as out simple benchmark model. The corresponding log-likelihood value comes out to be **-99.32**.

@@

A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions.

```{r, echo=FALSE, eval=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out

foreach (
  i=1:10,
  .combine=c,
  .options.future=list(seed=652643293)
) %dofuture% {
  mif2_out |> pfilter(Np=5)
} -> pf
logLik(pf) -> ll
print(logmeanexp(ll,se=TRUE))
```

The estimated log-likelihood is -205 with standard error of 3.14.

Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model. This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as to control the robustness. The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.

```{r, echo=FALSE, eval=FALSE}
bake(file="local_search.rds", {
  foreach(i=1:20,.combine=c,
    .options.future=list(seed=482947940)
  ) %dofuture% {
    mod |>
      mif2(
        Np=1000, Nmif=50,
        cooling.fraction.50=0.5,
        rw.sd = rw_sd(
           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
           a = 0.02, b = 0.02, c = 0.02, 
          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
        ),
        partrans = parameter_trans(
             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')),
        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs"),
        params = param_guess
      )
  }}) -> mifs_local

logLik(mifs_local) -> ll
print(logmeanexp(ll,se=TRUE))

@@

It estimated log likelihood of -134 which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
![Model parameters over iterations](./assets/local_search_1.png)
From the above plot, the likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

```{r}
pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```

![](./assets/local_search_2.png)
A scattered or cloud-like pattern indicates little to no apparent linear correlation.

@@

```{r}
set.seed(2062379496)

runif_design(
  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
    nseq=50
) -> guesses

mf1 <- mifs_local[[1]]
fixed_params <- coef(mif2_out)

bake(file="global_search_middle2.rds",
  dependson=guesses,{
    foreach(guess=iter(guesses,"row"), .combine=rbind,
      .options.future=list(seed=1270401374)
    ) %dofuture% {
      mf1 |>
        mif2(params=c(guess,fixed_params)) |>
        mif2(Nmif=100) -> mf
      replicate(
        10,
        mf |> pfilter(Np=500) |> logLik()
      ) |>
        logmeanexp(se=TRUE) -> ll
      mf |> coef() |> bind_rows() |>
        bind_cols(loglik=ll[1],loglik.se=ll[2])
    } -> results
    attr(results,"ncpu") <- nbrOfWorkers()
    results
  }) |>
  filter(is.finite(loglik)) -> results

bind_rows(results) |>
  filter(is.finite(loglik)) |>
  arrange(-loglik) |>
  write_csv("bird_params_middle2.csv")

```

```{r, echo=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_small.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

best_index <- which.max(results$loglik)
best_parameters <- results[best_index, ]
print(paste('The best parameters are:', toString(best_parameters)))
```

The best estimated log likelihood under global search by the use of GreatLakes is -320.4, which is a lot lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134). It is opposite to our expectation. One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters. The accuracy will be increased if we increase the parameter space.

As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric

Our report probes the potential of applying POMP to the field of ecological population dynamics. It ends with ARIMA(0,1,5) produces the highest log-likelihood with **-99**.  We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models. Drawing inspiration from the Lotka-Voltera equations, we introduce our take on the concept of an alternative prey hypothesis. Although our analysis is on the side of preliminary, it shows this approach can shed more light on the dynamic responses of ecosystems to changes in prey abundance.

@@

# Contribution
The contribution is not included in order to be blinded.
--------

### 2024-04-24 draft.Rmd
ready version

```{r, echo=FALSE, eval=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out
```

@@

```{r, echo=FALSE, eval=FALSE}
bake(file="local_search.rds", {
  foreach(i=1,.combine=c,
    .options.future=list(seed=482947940)
  ) %dofuture% {
    mod |>
      mif2(
        Np=1000, Nmif=50,
        cooling.fraction.50=0.5,
        rw.sd = rw_sd(
           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
           a = 0.02, b = 0.02, c = 0.02, 
          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
        ),
        partrans = parameter_trans(
             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')),
        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs"),
        params = param_guess
      )
  }}) -> mifs_local

@@

The likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

@@

# Results

@@

Our report probes the potential of applying POMP to the field of ecological population dynamics. We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models. Drawing inspiration from the Lotka-Voltera equations, we introduce our take on the concept of an alternative prey hypothesis. Although our analysis is on the side of preliminary, it shows this approach can shed more light on the dynamic responses of ecosystems to changes in prey abundance.

@@

- Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
- Heyuan: 
- Sizhuang:

>>>>

```{r, echo=FALSE, eval=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out

foreach (
  i=1:10,
  .combine=c,
  .options.future=list(seed=652643293)
) %dofuture% {
  mif2_out |> pfilter(Np=5)
} -> pf
logLik(pf) -> ll
print(logmeanexp(ll,se=TRUE))
```

The estimated log-likelihood is -205 with standard error of 3.14.

@@

```{r, echo=FALSE, eval=FALSE}
bake(file="local_search.rds", {
  foreach(i=1:20,.combine=c,
    .options.future=list(seed=482947940)
  ) %dofuture% {
    mod |>
      mif2(
        Np=1000, Nmif=50,
        cooling.fraction.50=0.5,
        rw.sd = rw_sd(
           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
           a = 0.02, b = 0.02, c = 0.02, 
          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
        ),
        partrans = parameter_trans(
             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')),
        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs"),
        params = param_guess
      )
  }}) -> mifs_local

logLik(mifs_local) -> ll
print(logmeanexp(ll,se=TRUE))

@@

It estimated log likelihood of -134 which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
![Model parameters over iterations](./assets/local_search_1.png)
From the above plot, the likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

```{r}
pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```

![](./assets/local_search_2.png)
A scattered or cloud-like pattern indicates little to no apparent linear correlation.

@@

```{r}
set.seed(2062379496)

runif_design(
  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
    nseq=50
) -> guesses

mf1 <- mifs_local[[1]]
fixed_params <- coef(mif2_out)

bake(file="global_search_middle2.rds",
  dependson=guesses,{
    foreach(guess=iter(guesses,"row"), .combine=rbind,
      .options.future=list(seed=1270401374)
    ) %dofuture% {
      mf1 |>
        mif2(params=c(guess,fixed_params)) |>
        mif2(Nmif=100) -> mf
      replicate(
        10,
        mf |> pfilter(Np=500) |> logLik()
      ) |>
        logmeanexp(se=TRUE) -> ll
      mf |> coef() |> bind_rows() |>
        bind_cols(loglik=ll[1],loglik.se=ll[2])
    } -> results
    attr(results,"ncpu") <- nbrOfWorkers()
    results
  }) |>
  filter(is.finite(loglik)) -> results

bind_rows(results) |>
  filter(is.finite(loglik)) |>
  arrange(-loglik) |>
  write_csv("bird_params_middle2.csv")

```

```{r, echo=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

best_index <- which.max(results$loglik)
best_parameters <- results[best_index, ]
print(paste('The best parameters are:', toString(best_parameters)))
```

The best estimated log likelihood under global search by the use of GreatLakes is -176.3, which is a lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134). It is opposite to our expectation. One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters. The accuracy will be increased if we increase the parameter space.

As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric

@@

Our report probes the potential of applying POMP to the field of ecological population dynamics. It ends with ARIMA(0,1,5) produces the highest log-likelihood with **-99**.  We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models. Drawing inspiration from the Lotka-Voltera equations, we introduce our take on the concept of an alternative prey hypothesis. Although our analysis is on the side of preliminary, it shows this approach can shed more light on the dynamic responses of ecosystems to changes in prey abundance.

@@

- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections, ARMA. 
- Heyuan: Wrote the interpretation of result, refine the ARIMA, do the POMP, local search and global search. 
- Sizhuang: Helped with POMP, local search and global search and created visualizations of some of the POMP results
Professor and GSI helped us a lot in building POMP model.
--------

### 2024-04-24 unblinded.Rmd
ready version

```{r, echo=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

@@

The best estimated log likelihood under global search by the use of GreatLakes is -176.3, which is a lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134). It is opposite to our expectation. One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters. The accuracy will be increased if we increase the parameter space.

@@

- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections, ARMA. 
- Heyuan: Wrote the interpretation of result, refine the ARIMA, do the POMP, local search and global search. 
- Sizhuang: Helped with POMP, local search and global search and created visualizations of some of the POMP results
Professor and GSI helped us a lot in building POMP model.
--------

>>>>

```{r, echo=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_small.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

@@

The best estimated log likelihood under global search by the use of GreatLakes is -320.4, which is a lot lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134). It is opposite to our expectation. One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters. The accuracy will be increased if we increase the parameter space.

@@

- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections. 
- Heyuan: Wrote the interpretation of result, do the POMP, local search and global search. 
- Sizhuang: Helped with POMP, local search and global search and created visualizations of some of the POMP results
Professor and GSI helped us a lot in building POMP model.
--------

### 2024-04-24 workspace/sizhuang.Rmd
ready version

>>>>

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Load data
data <- read.csv("/Users/sizhuang/Desktop/bird_params_small.csv")

# Reshape data to long format
data_long <- data %>%
  pivot_longer(
    cols = -c(loglik, loglik.se), # Exclude loglik and loglik.se from reshaping
    names_to = "variable",
    values_to = "value"
  )

# Calculate the threshold for the red line
threshold <- max(data$loglik) - 0.5 * qchisq(p = 0.95, df = 1)

# Define the number of variables per plot
vars_per_plot <- 5  # Adjust this as needed to fit your specific requirements
total_vars <- length(unique(data_long$variable))
total_plots <- ceiling(total_vars / vars_per_plot)

# Create and display/save multiple plots
plots <- list()
for (i in 1:total_plots) {
  subset_vars <- unique(data_long$variable)[((i-1) * vars_per_plot + 1):(i * vars_per_plot)]
  subset_data <- data_long %>% filter(variable %in% subset_vars)

  # Adjusted for more columns to expand vertical space
  ncol_facet = min(3, length(subset_vars))  # Use 3 columns or less based on available variables
  nrow_facet = ceiling(length(subset_vars) / ncol_facet)  # Calculate number of rows based on number of columns

  plots[[i]] <- ggplot(subset_data, aes(x = value, y = loglik)) +
    geom_point(alpha = 0.3) +
    geom_hline(yintercept = threshold, color = "red", linetype = "dashed", size = 1) +
    facet_wrap(~ variable, scales = "free_x", ncol = ncol_facet, nrow = nrow_facet) +
    labs(title = paste("Variables vs Loglik - Plot", i),
         x = "Value",
         y = "Loglik") +
    theme_minimal() +
    theme(strip.text.x = element_text(size = 12, face = "bold"),
          legend.position = "none",
          axis.title.y = element_text(size = 14),
          axis.title.x = element_text(size = 14),
          panel.spacing = unit(1.5, "lines"))

  # Display or save plots
  print(plots[[i]])
  # Optionally save each plot
  # ggsave(paste("Plot_", i, ".png", sep = ""), plot, width = 20, height = 8 * nrow_facet, units = "in")
}
